---
title: "Lab 5"
author: "David Fainstein"
date: "October 21, 2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE) 
```

1. Create a RStudio Project on your machine.

* Within the Project create a folder for *data* and *scripts*.
* Download the following three datasets into the project *data* folder.
  + *sesame13.sav*
  + *star.csv*
  + *ais.xlsx*

2. Read in the following three datasets, using `{here}` and `{rio}`:
  * *sesame13.sav*
  * *star.csv*
  * *ais.xlsx*
  
**Hint**: For the *ais.xlsx* dataset look at the `skip` argument within the `rio` help documentation.
  
```{r import}

library(rio)
library(tidyverse)
library(here)
library(openxlsx)
library(foreign)

sesame <- import(here("data", "sesame13.sav"))
ais <- import(here("data", "ais.xlsx"), skip = 15)
star <- import(here("data", "star.csv"))

```

3. Using the `ais` data, compute the average red blood cell count and average bmi by sport. Output these data as SPSS and EXCEL files. 

```{r q2}

ais %>%
  select("rcc", "bmi") %>%
  summarize(mean_rcc = mean(rcc), mean_bmi = mean(bmi)) %>% 
  export(here("data", "means.xlsx")) 

ais %>%
  select("rcc", "bmi") %>%
  summarize(mean_rcc = mean(rcc), mean_bmi = mean(bmi)) %>% 
  export(here("data", "means2.sav")) 
```

4. Use the `sesame` dataset to answer the following question: Was the average female age higher in schools or at home? 

```{r q3}

sesame %>% 
  select("sex", "age", "setting") %>% 
  filter(sex == "2") %>% 
  group_by(setting) %>% 
  summarize(mean_age_schools = mean(age))
  
```
Note: could not tell which sex (1,2) represented which gender, nor which setting (0,1) represented home or school. With that information, then the question is answerable with the data made available


5. First, how many rows and columns are in the *star* dataset? Next, remove outliers using a really poor method, just for practice, by eliminating students whose math (*tmathss*) scores were more than three standard deviations **above or below** the corresponding mean. How many rows are in the dataset now?

```{r q4}
nrow(star)
ncol(star)

sdcrit <- sd(star$tmathss)*3
meanvalue <- mean(star$tmathss)
lowerbound <- meanvalue-sdcrit
upperbound <- meanvalue+sdcrit

star2 <- star %>% 
  filter(tmathss > lowerbound, tmathss < upperbound)
  

nrow(star2)
ncol(star2)
```
We went from 5748 rows to 5743 rows. That's not very many outliers removed!


6. Use the `star` dataset to compute standardized math and reading scores; name these variables *tmathss* and *treadss*. This means, for each of these variables, subtract the mean from each observation and divide by the standard deviation: $x_s = \frac{x_i - \bar{X}}{sd(X)}$. 
  + Check that your computations were correct by computing the mean and standard deviation of each variable (they should be 0 and 1). 
  + Compute the mean of the standardized variable for all *sex*/*frl* combinations (I'm asking you to extend what you know here, we haven't talked explicitly about how to do this yet, but you have seen examples). 
  + What do you make of these findings? Do you see an effect by *sex*? An *FRL* effect? Is there evidence of an interaction (i.e., that the effect of *FRL* is greater for boys versus girls)?

```{r q5}

sdvalue <- sd(star$tmathss)
meanvalue <- mean(star$tmathss)
sdvalue_r <- sd(star$treadss)
meanvalue_r <- mean(star$treadss)


star3 <- star %>% 
  mutate(tmathss = (tmathss-meanvalue)/sdvalue) %>% 
  mutate(treadss = (treadss-meanvalue_r)/sdvalue_r)# %>% 
#  summarize(mean_r = mean(treadss), sd_r = sd(treadss), mean_m = mean(tmathss), sd_m = sd(tmathss))

star4 <- star3 %>% 
  select("sex", "frl", "tmathss", "treadss") %>% 
  group_by(sex, frl) %>% 
  summarize(mean_m = mean(tmathss), mean_r = mean(treadss))



```
Interestingly, the mean of means is not zero for the standardized scores; though, it is exceptionally small. I'm guessing this has something to do with the type of numeric value (e.g., double)?

It looks as though there is an interaction with free-reduced price lunch and students scores because student's of either gender, as coded, scored below the mean on average if they were receiving FRL. 

Please submit you .Rmd file, as well as the rendered .html file.